---
title: "Hands-on-Exercise 3.2:  2nd Order Spatial Point Patterns Analysis Methods"
author: "Gerald Lim"

execute:
  warning: false
  freeze: false
date: "`r Sys.Date()`"
---

## **5.1 Overview**

Spatial Point Pattern Analysis is the evaluation of the pattern or distribution, of a set of points on a surface. The point can be location of:

-   events such as crime, traffic accident and disease onset, or

-   business services (coffee and fastfood outlets) or facilities such as childcare and eldercare.

Using appropriate functions of [spatstat](https://cran.r-project.org/web/packages/spatstat/), this hands-on exercise aims to discover the spatial point processes of childecare centres in Singapore.

The specific questions we would like to answer are as follows:

-   are the childcare centres in Singapore randomly distributed throughout the country?

-   if the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

## **5.2 The data**

To provide answers to the questions above, three data sets will be used. They are:

-   `CHILDCARE`, a point feature data providing both location and attribute information of childcare centres. It was downloaded from Data.gov.sg and is in geojson format.

-   `MP14_SUBZONE_WEB_PL`, a polygon feature data providing information of URA 2014 Master Plan Planning Subzone boundary data. It is in ESRI shapefile format. This data set was also downloaded from Data.gov.sg.

-   `CostalOutline`, a polygon feature data showing the national boundary of Singapore. It is provided by SLA and is in ESRI shapefile format.

## **5.3 Installing and Loading the R packages**

In this hands-on exercise, five R packages will be used, they are:

-   [**sf**](https://r-spatial.github.io/sf/), a relatively new R package specially designed to import, manage and process vector-based geospatial data in R.

-   [**spatstat**](https://spatstat.org/), which has a wide range of useful functions for point pattern analysis. In this hands-on exercise, it will be used to perform 1st- and 2nd-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.

-   [**raster**](https://cran.r-project.org/web/packages/raster/) which reads, writes, manipulates, analyses and model of gridded spatial data (i.e. raster). In this hands-on exercise, it will be used to convert image output generate by spatstat into raster format.

-   [**maptools**](https://cran.r-project.org/web/packages/maptools/) which provides a set of tools for manipulating geographic data. In this hands-on exercise, we mainly use it to convert *Spatial* objects into *ppp* format of **spatstat**.

-   [**tmap**](https://cran.r-project.org/web/packages/tmap/) which provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [leaflet](https://leafletjs.com/) API.

Use the code chunk below to install and launch the five R packages.

```{r}
#| eval: false
install.packages("maptools", repos = "https://packagemanager.posit.co/cran/2023-10-13")
```

```{r}
pacman::p_load(maptools, sf, raster, spatstat, tmap, tidyverse)
```

## **5.4 Spatial Data Wrangling**

### **5.4.1 Importing the spatial data**

In this section, [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package will be used to import these three geospatial data sets into R.

```{r}
childcare_sf <- st_read("data/child-care-services-geojson.geojson") %>%
  st_transform(crs = 3414)
```

```{r}
sg_sf <- st_read(dsn = "data/", 
                layer = "CostalOutline")
```

```{r}
mpsz_sf <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

Before we can use these data for analysis, it is important for us to ensure that they are projected in same projection system.

> DIY: Using the appropriate sf function you learned in Hands-on Exercise 2, retrieve the referencing system information of these geospatial data.

```{r}
st_crs(childcare_sf)
```

```{r}
st_crs(mpsz_sf)
```

```{r}
st_crs(childcare_sf)
```

Notice that except `childcare_sf`, both `mpsz_sf` and `sg_sf` do not have proper crs information.

> DIY: Using the method you learned in Lesson 2, assign the correct crs to mpsz_sf and sg_sf simple feature data frames.

The cr information isn’t appropriate. `childcare_sf`  and 'sg_sf' are in WGS84, while the 'other two are'mpsz_sf' is in SVY21

```{r}
mpsz_sf <- st_transform(mpsz_sf, crs= 3414)
st_crs(mpsz_sf)
```

### **5.4.2 Mapping the geospatial data sets**

After checking the referencing system of each geospatial data data frame, it is also useful for us to plot a map to show their spatial patterns.

> DIY: Using the mapping methods you learned in Hands-on Exercise 3, prepare a map as shown below.

Let's plot a map to show their spatial patterns for each dataframe

childcare_sf consists of spatial points, so it cannot accept tm_fill/tm_borders/tm_polygons. So we have to use tm_dots()

```{r}
invalid_mpsz_sf <- which(!st_is_valid(mpsz_sf))
invalid_sg_sf <- which(!st_is_valid(sg_sf))
invalid_childcare_sf <- which(!st_is_valid(childcare_sf))

# Print the indices of the invalid geometries
print(invalid_mpsz_sf)
print(invalid_sg_sf)
print(invalid_childcare_sf)
```

Make sg_sf and mpsz_sf valid

```{r}
sg_sf <- st_make_valid(sg_sf)
mpsz_sf <- st_make_valid(mpsz_sf)
```

```{r}
# Now, try plotting again
tm_shape(sg_sf) +
  tm_polygons() +
tm_shape(mpsz_sf) +
  tm_polygons() +
tm_shape(childcare_sf)+
  tm_dots()
```

Notice that all the geospatial layers are within the same map extend. This shows that their referencing system and coordinate values are referred to similar spatial context. This is very important in any geospatial analysis.

Alternatively, we can also prepare a pin map by using the code chunk below.

```{r}
tmap_mode('view')
```

```{r}
tm_shape(childcare_sf)+
  tm_dots()
```

```{r}
tmap_mode('plot')
```

Notice that at the interactive mode, **tmap** is using [**leaflet for R**](https://rstudio.github.io/leaflet/) API. The advantage of this interactive pin map is it allows us to navigate and zoom around the map freely. We can also query the information of each simple feature (i.e. the point) by clicking of them. Last but not least, you can also change the background of the internet map layer. Currently, three internet map layers are provided. They are: ESRI.WorldGrayCanvas, OpenStreetMap, and ESRI.WorldTopoMap. The default is ESRI.WorldGrayCanvas.

> Reminder: Always remember to switch back to plot mode after the interactive map. This is because, each interactive mode will consume a connection. You should also avoid displaying ecessive numbers of interactive maps (i.e. not more than 10) in one RMarkdown document when publish on Netlify.

## **5.5 Geospatial Data wrangling**

Although simple feature data frame is gaining popularity again sp’s Spatial\* classes, there are, however, many geospatial analysis packages require the input geospatial data in sp’s Spatial\* classes. In this section, you will learn how to convert simple feature data frame to sp’s Spatial\* class.

### **5.5.1 Converting sf data frames to sp’s Spatial\* class**

The code chunk below uses [*as_Spatial()*](https://r4gdsa.netlify.app/chap05) of **sf** package to convert the three geospatial data from simple feature data frame to sp’s Spatial\* class.

```{r}
childcare <- as_Spatial(childcare_sf)
mpsz <- as_Spatial(mpsz_sf)
sg <- as_Spatial(sg_sf)
```

DIY: Using appropriate function, display the information of these three Spatial\* classes as shown below.

```{r}
childcare
```

```{r}
mpsz
```

```{r}
sg
```

Notice that the geospatial data have been converted into their respective sp’s Spatial\* classes now.

### **5.5.2 Converting the Spatial\* class into generic sp format**

**spatstat** requires the analytical data in ***ppp*** object form. There is no direct way to convert a Spatial\* classes into ***ppp*** object. We need to convert the ***Spatial*** **classes**\* into ***Spatial*** object first.

The codes chunk below converts the Spatial\* classes into generic sp objects.

```{r}
childcare_sp <- as(childcare, "SpatialPoints")
sg_sp <- as(sg, "SpatialPolygons")
```

Next, you should display the sp objects properties as shown below.

```{r}
childcare_sp
```

```{r}
sg_sp
```

### **5.5.3 Converting the generic sp format into spatstat’s ppp format**

Now, we will use *as.ppp()* function of **spatstat** to convert the spatial data into **spatstat**’s ***ppp*** object format.

```{r}
library(spatstat)

# Assuming childcare_sp is a SpatialPoints object
coords <- coordinates(childcare_sp)

# Create a rectangular window covering the entire extent of the points
window <- owin(xrange = range(coords[, 1]), yrange = range(coords[, 2]))

# Create a ppp object with the adjusted window
childcare_ppp <- ppp(coords[, 1], coords[, 2], window = window)

childcare_ppp
```

```         
Planar point pattern: 1545 points window: rectangle = [11203.01, 45404.24] x [25667.6, 49300.88] units
```

Now, let us plot ***childcare_ppp*** and examine the difference.

```{r}
plot(childcare_ppp)
```

You can take a quick look at the summary statistics of the newly created ppp object by using the code chunk below.

```{r}
summary(childcare_ppp)
```

Notice the warning message about duplicates. In spatial point patterns analysis an issue of significant is the presence of duplicates. The statistical methodology used for spatial point patterns processes is based largely on the assumption that process are *simple*, that is, that the points cannot be coincident.

### **5.5.4 Handling duplicated points**

We can check the duplication in a ***ppp*** object by using the code chunk below.

```{r}
any(duplicated(childcare_ppp))
```

To count the number of co-indicence point, we will use the *multiplicity()* function as shown in the code chunk below.

```{r}
multiplicity(childcare_ppp)
```

If we want to know how many locations have more than one point event, we can use the code chunk below.

```{r}
sum(multiplicity(childcare_ppp) > 1)
```

The output shows that there are 128 duplicated point events.

To view the locations of these duplicate point events, we will plot `childcare` data by using the code chunk below.

```{r}
tmap_mode('view')
```

```{r}
tm_shape(childcare) +
  tm_dots(alpha=0.4, 
          size=0.05)
```

```{r}
tmap_mode('plot')
```

There are three ways to overcome this problem. The easiest way is to delete the duplicates. But, that will also mean that some useful point events will be lost.

The second solution is use *jittering*, which will add a small perturbation to the duplicate points so that they do not occupy the exact same space.

The third solution is to make each point “unique” and then attach the duplicates of the points to the patterns as **marks**, as attributes of the points. Then you would need analytical techniques that take into account these marks.

The code chunk below implements the jittering approach.

```{r}
childcare_ppp_jit <- rjitter(childcare_ppp, 
                             retry=TRUE, 
                             nsim=1, 
                             drop=TRUE)
```

```{r}
any(duplicated(childcare_ppp_jit))
```

### **5.5.5 Creating *owin* object**

When analysing spatial point patterns, it is a good practice to confine the analysis with a geographical area like Singapore boundary. In spatstat, an object called owin is specially designed to represent this polygonal region.

The code chunk below is used to covert sg SpatialPolygon object into owin object of spatstat.

```{r}
sg_sp
```

```{r}
# Convert the SpatialPolygons object sg_sp to owin
sg_owin <- as.owin(sg_sf)
```

The output object can be displayed by using plot() function

```{r}

# Plot the map
plot(sg_owin, main = "sg_owin")

```

Summary() function of Base R

```{r}
summary(sg_owin)
```

### **5.5.6 Combining point events object and owin object**

In this last step of geospatial data wrangling, we will extract childcare events that are located within Singapore by using the code chunk below.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
```

The output object combined both the point and polygon feature in one ppp object class as shown below.

```{r}
summary(childcareSG_ppp)
```

Plot the newly derived childcareSG_ppp

```{r}
plot(childcareSG_ppp)
```

#### 5.5.6.1 Extracting study area

The code chunk below will be used to extract the target planning areas.

```{r}
pg = mpsz[mpsz@data$PLN_AREA_N == "PUNGGOL",]
tm = mpsz[mpsz@data$PLN_AREA_N == "TAMPINES",]
ck = mpsz[mpsz@data$PLN_AREA_N == "CHOA CHU KANG",]
jw = mpsz[mpsz@data$PLN_AREA_N == "JURONG WEST",]
```

Plotting target planning areas

```{r}
par(mfrow=c(2,2))
plot(pg, main = "Punggol")
plot(tm, main = "Tampines")
plot(ck, main = "Choa Chu Kang")
plot(jw, main = "Jurong West")
```

#### 5.5.6.2 Converting the spatial point data frame into generic sp format

Next, we will convert these SpatialPolygonsDataFrame layers into generic spatialpolygons layers.

```{r}
pg_sp = as(pg, "SpatialPolygons")
tm_sp = as(tm, "SpatialPolygons")
ck_sp = as(ck, "SpatialPolygons")
jw_sp = as(jw, "SpatialPolygons")
```

#### 5.5.6.3 Creating ***owin*** object

Now, we will convert these SpatialPolygons objects into owin objects that is required by **spatstat**.

```{r}
library(spatstat)

# Assuming pg_sp is your SpatialPolygons object
pg_coords <- coordinates(pg_sp)
pg_xrange <- range(pg_coords[, 1])
pg_yrange <- range(pg_coords[, 2])
pg_owin <- owin(c(pg_xrange[1], pg_xrange[2]), c(pg_yrange[1], pg_yrange[2]))

# Repeat the above process for tm_sp, ck_sp, and jw_sp

# Assuming tm_sp is your SpatialPolygons object
tm_coords <- coordinates(tm_sp)
tm_xrange <- range(tm_coords[, 1])
tm_yrange <- range(tm_coords[, 2])
tm_owin <- owin(c(tm_xrange[1], tm_xrange[2]), c(tm_yrange[1], tm_yrange[2]))

# Assuming ck_sp is your SpatialPolygons object
ck_coords <- coordinates(ck_sp)
ck_xrange <- range(ck_coords[, 1])
ck_yrange <- range(ck_coords[, 2])
ck_owin <- owin(c(ck_xrange[1], ck_xrange[2]), c(ck_yrange[1], ck_yrange[2]))

# Assuming jw_sp is your SpatialPolygons object
jw_coords <- coordinates(jw_sp)
jw_xrange <- range(jw_coords[, 1])
jw_yrange <- range(jw_coords[, 2])
jw_owin <- owin(c(jw_xrange[1], jw_xrange[2]), c(jw_yrange[1], jw_yrange[2]))

# Now pg_owin, tm_owin, ck_owin, jw_owin are owin objects representing the bounding boxes
```

#### 5.5.6.4 Combining childcare points and the study area

By using the code chunk below, we are able to extract childcare that is within the specific region to do our analysis later on.

```{r}
childcare_pg_ppp = childcare_ppp_jit[pg_owin]
childcare_tm_ppp = childcare_ppp_jit[tm_owin]
childcare_ck_ppp = childcare_ppp_jit[ck_owin]
childcare_jw_ppp = childcare_ppp_jit[jw_owin]
```

Next, *rescale()* function is used to trasnform the unit of measurement from metre to kilometre.

```{r}
childcare_pg_ppp.km = rescale(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale(childcare_jw_ppp, 1000, "km")
```

The code chunk below is used to plot these four study areas and the locations of the childcare centres.

```{r}
par(mfrow=c(2,2))
plot(childcare_pg_ppp.km, main="Punggol")
plot(childcare_tm_ppp.km, main="Tampines")
plot(childcare_ck_ppp.km, main="Choa Chu Kang")
plot(childcare_jw_ppp.km, main="Jurong West")
```

## **5.6 Second-order Spatial Point Patterns Analysis**

## **5.7 Analysing Spatial Point Process Using G-Function**

The G function measures the distribution of the distances from an arbitrary event to its nearest event. In this section, you will learn how to compute G-function estimation by using [*Gest()*](https://rdrr.io/cran/spatstat/man/Gest.html) of **spatstat** package. You will also learn how to perform monta carlo simulation test using [*envelope()*](https://rdrr.io/cran/spatstat/man/envelope.html) of **spatstat** package.

### **5.7.1 Choa Chu Kang planning area**

#### 5.7.1.1 Computing G-function estimation

The code chunk below is used to compute G-function using *Gest()* of **spatat** package.

```{r}
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim=c(0,500))
```

#### 5.7.1.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with G-fucntion

```{r}
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
```

```{r}
plot(G_CK.csr)
```

### **5.7.2 Tampines planning area**

#### 5.7.2.1 Computing G-function estimation

```{r}
G_tm = Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

#### 5.7.2.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Tampines are randomly distributed.

H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected is p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
```

```{r}
plot(G_tm.csr)
```

## **5.8 Analysing Spatial Point Process Using F-Function**

The F function estimates the empty space function F(r) or its hazard rate h(r) from a point pattern in a window of arbitrary shape. In this section, you will learn how to compute F-function estimation by using [*Fest()*](https://rdrr.io/cran/spatstat/man/Fest.html) of **spatstat** package. You will also learn how to perform monta carlo simulation test using [*envelope()*](https://rdrr.io/cran/spatstat/man/envelope.html) of **spatstat** package.

### **5.8.1 Choa Chu Kang planning area**

#### 5.8.1.1 Computing F-function estimation

The code chunk below is used to compute F-function using *Fest()* of **spatat** package.

```{r}
F_CK = Fest(childcare_ck_ppp)
plot(F_CK)
```

### **5.8.2 Performing Complete Spatial Randomness Test**

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with F-fucntion

```{r}
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
```

```{r}
plot(F_CK.csr)
```

### **5.8.3 Tampines planning area**

#### 5.8.3.1 Computing F-function estimation

Monte Carlo test with F-fucntion

```{r}
F_tm = Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

#### 5.8.3.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Tampines are randomly distributed.

H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected is p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
```

```{r}
plot(F_tm.csr)
```

## **5.9 Analysing Spatial Point Process Using K-Function**

K-function measures the number of events found up to a given distance of any particular event. In this section, you will learn how to compute K-function estimates by using [*Kest()*](https://rdrr.io/cran/spatstat/man/Kest.html) of **spatstat** package. You will also learn how to perform monta carlo simulation test using *envelope()* of spatstat package.

### **5.9.1 Choa Chu Kang planning area**

#### 5.9.1.1 Computing K-fucntion estimate

```{r}
K_ck = Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

#### 5.9.1.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

### **5.9.2 Tampines planning area**

#### 5.9.2.1 Computing K-fucntion estimation

```{r}
K_tm = Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . -r ~ r, 
     ylab= "K(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

#### 5.9.2.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Tampines are randomly distributed.

H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_tm.csr, . - r ~ r, 
     xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

## **5.10 Analysing Spatial Point Process Using L-Function**

In this section, you will learn how to compute L-function estimation by using [*Lest()*](https://rdrr.io/cran/spatstat/man/Lest.html) of **spatstat** package. You will also learn how to perform monta carlo simulation test using *envelope()* of spatstat package.

### **5.10.1 Choa Chu Kang planning area**

#### 5.10.1.1 Computing L Fucntion estimation

```{r}
L_ck = Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

#### 5.10.1.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value if smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```

### **5.10.2 Tampines planning area**

#### 5.10.2.1 Computing L-fucntion estimate

```{r}
L_tm = Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

#### 5.10.2.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Tampines are randomly distributed.

H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below will be used to perform the hypothesis testing.

```{r}
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

Then, plot the model output by using the code chun below.

```{r}
plot(L_tm.csr, . - r ~ r, 
     xlab="d", ylab="L(d)-r", xlim=c(0,500))
```
